Sketch Recognition using Particle Swarm Optimizations

### 2009-06-10 trunk/paper2/Paper2.tex
after add

Recognition based on AI methods as applied in \cite{SketchRead2007} has the principle problem of computational complexity of the matching between shape description and geometric primitives produced in the segmentation process. Natural sketch are difficult to be segmented and grouped to create reliable fragment of geometric primitives. This is partially due to the increase of noise and other phenomena such as overlapped, over traced strokes and sharing of strokes between different symbols. Since these systems are dependable on the segmentation results, they either perform recognition using undependable segmentation results or consider different ways to segment the input sketch. This is because they assume that each stroke plays a role in the structure of the shape. 
%The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods.  
%Gestures are pen strokes drawn by the user that can immediately recognized by the system.
%role of which each stroke plays in the structure of the sketched symbol is used to determine shapes 
%. Describing the role  how user draws individual strokes  properties that describe how users draw individual strokes in drawing the symbol.  to. The identification of individual stroke is used to determine The symbols are recognized by identifying the individual strokes.

@@

 %In general, the appearance based methods use image processing techniques to recognize sketches that the users draw. 
%Sketch recognition can be based on: strokes, global shape properties, and appearance \cite{Oltmans07}. The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods. Most stroke based systems either represent symbols as gestures or as a hierarchical shape. Gestures are pen strokes drawn by the user that can immediately recognized by the system. A Hierarchical shape descriptions was widely used in sketch systems \cite{HierarchicalParsing7,Ladder30,AlvaradoFreedom42,SketchRead2007}. Symbols are represented using hierarchical levels of simpler geometric shapes or primitives and spatial relations between them. Systems that employ such representation must first pre-process strokes to break them into geometric primitives. Recognition can then be treated as sub graph matching problem.

 %In global shape properties approach, systems investigate general shape properties and its underlying strokes. Using shape properties (i.e. ratio of area of convex hull to area of shape) relaxes the assumption and role of each stroke in the symbol but it does not truly represent the appearance of the symbol\cite{DiagramOfflineConvexHull}. Unlike appearance based methods where individual strokes are disregarded and they focus on the overall appearance of shapes. In general, the appearance based methods use image processing techniques to recognize sketches that the users draw.

 %Methods based on gesture and global shape properties produce systems that are dependent on assumptions about how strokes can be grouped. These systems assume that users will finish drawing a shape before moving to the next one \cite{Cali63,geometrydomain49}. This let them to temporally group strokes and then recognizes them based on their strokes or global shape properties. \cite{AlvaradoDigital} argued that this assumption does not hold in natural sketches.

 %Recognition based on strokes using hierarchical shape description is applied in \cite{SketchRead2007}. The principle problem in these systems is the computational complexity of the matching between shape description and geometric primitives produced in the segmentation process. Natural sketch are difficult to segment and group to create reliable fragment of geometric primitives. This is partially due to the increase of noise and other phenomena such as overlapped, over traced strokes and sharing of strokes between different symbols. Since these systems are dependable on the segmentation results, they either perform recognition using undependable segmentation results or consider different ways to segment the input sketch.

%In this paper we choose to implement a hybrid method between shape property and stroke based approaches. Our method tries to avoid many of the problems faced by approaches based on individual strokes or global shape properties while trying to gain the additional information that the stroke based system use. Firstly, the system uses particle swarm optimization (PSO) to optimally segment user's strokes. Users can draw symbols using any number of strokes in any order they like. The system segments each stroke using two different PSO algorithms. The first segmentation algorithm is based on dividing the digital curve into polygons\cite{PolygonApproximationPSO}. Our algorithm enhances its performance by adding curvature and other local information while segmenting strokes to achieve better results. To handle complex strokes, our system uses a second segmentation PSO algorithm to segment strokes into a set of lines and circular arcs. After segmentation, a feature vector composed of different statistical, geometrical and composite features is deduced. The final classification step uses a SVM classifier to classify symbols into one of the previously trained classes based on the features vector computed.

 %The preprocessing phase captures user input stroke points and collects basic information about the stroke. In the segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier.

%This paper introduces a new sketch recognition system, which uses  % uses the computed feature vector to classify the segments into one of the previously trained classes.
%Sections \ref{Prepross}, \ref{seg} and \ref{sec:Recognition} provide details on preprocessing, segmentation and recognition process respectively.
 The remaining of the paper is as follows; the literature review on stroke segmentation is presented in Section \ref{sec:review}. Section \ref{sec:ParticleSwarmAlgorithm} introduces Particle Swarm Optimization. Section \ref{Sysdisc} describes the proposed system in details. Section \ref{sec:Experiments} describes the results obtained from the experiments that were preformed. Finally, the conclusion and future work is presented in Section \ref{ConclusionandFutureWork}.

@@

\begin{equation}
%\[
p_{ij}=p_{ij}+v_{ij},
%\
\label{eq:Swarm1}
\end{equation}
%where $p_{ij}$ represent the $jth$ particle in the $ith$ agent and $v_{ij}$ is the velocity of the $jth$ particle in the $ith$ agent.
where $p_{ij}$ represent the $j$th dimension in the $i$th particle and $v_{ij}$ is the velocity of the $j$th dimension in the $i$th particle.
 %Equation [\ref{eq:Swarm}] shows how velocity and direction of each particle are computed
 \begin{equation}
v_{ij}  = v_{ij} \omega + c_1 r_1 (lbest_{ij}  - p_{ij} ) + c_2 r_2 (gbest_{ij}  - p_{ij} ),
\label{eq:Swarm}
\end{equation}
 where $\omega$ is the inertia weight parameter which controls the tradeoff between exploration and exploitation,  $lbest_{ij}$ is the local best particle, $gbest_{ij}$ is the global best particle, $r_1$ \& $r_2$ are random variables and $c_1$ \& $c_2$ are the swarm acceleration parameters.

@@

\subsection{Preprocessing}
\label{Prepross}% Agar \cite{polygonfeedback31} mentions that the pointing device (for example the mouse) sampling rate is the reason for this phenomena.
 Speed and time difference information were widely used in sketch recognition systems \cite{earlyprocess}. Figure \ref{fig:orignalStroke} shows an example of an input stroke drawn by users.  Figure \ref{fig:speed2Distance} shows the graph of time difference and the speed for the stroke in Figure \ref{fig:orignalStroke}. It was noted that time difference provides more distinct maxima than the speed information. Agar et al. \cite{polygonfeedback31} mentions that the pointing device (e.g., mouse) sampling rate is the reason for this phenomenon. The points are sampled at regular time intervals as long as the pointing device is moving. Therefore, there are no new detected points when the pointing device is stationary. This leads to a nearly constant time difference between samples while the pen is moving and large difference while the pen is stationary (Figure \ref{fig:timediff}). Contrary to time difference information, the speed information has a lot of noise in the data because usually  users draw with variant speeds (Figure \ref{fig:speed2Distance}).  %Similarly, direction information is used as it provides better distinctive maxima for the corners than the curvature information \cite{meanshift10}.% Figure \ref{fig:curvatures} shows the direction and curvature graphs for the stroke drawn in Figure \ref{fig:orignalStroke}

@@

 \begin{figure}
	\centering
			\subfigure[Speed versus Displacement Graph]{	\includegraphics[scale=0.4]{images/vel3.jpg}}
			\hfill
			\subfigure[ Time Difference  versus Displacement Graph] {\label{fig:timediff}\includegraphics[scale=0.4]{images/td3.jpg}}
	\caption{Speed and Time Difference Graphs}  a) speed graph has lot of noise in the data with respect to b) time difference.   The black points denote the location of the possible dominant points $P_{pd}$. Black rectangles represent regions $R_i$ that in a) Are lower than the threshold. b) Are above the threshold. 
	\label{fig:speed2Distance}
\end{figure}
%is calculated as the angle between two vectors and  where $Q_i=\overrightarrow {P_{i - 1} P_i }$ is the vector from the point $P_{i - 1}$ to point $P_i$ 
Dominant points are characterized by low speed values, high curvature and direction values. However, the dominant points from the curvature information may not be the same points as the dominant points extracted from the speed information. Hence, we compute time difference, direction, speed and curvature of each point along a stroke to ensure that no dominant point is overlooked in this phase. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ calculated as the angle between vectors $\overrightarrow {P_{i - 1} P_i }$ and the $x-axis$ and the curvature is considered as the change in direction $d$ with respect to length $s$ i.e. $c= \Delta d/\Delta s$.

@@

\subsection{Segmentation}
\label{seg}
In the segmentation stage, the input stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{ellipsefit}. The ellipse fitting step mainly helps to prevent the system from over segmenting the stroke into multiple lines or curves if the input stroke is an ellipse. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two further segmentation algorithms that divide the stroke to either lines or lines and curves. Hence we generate two different segmentations; the system chooses the segmentation with the minimum error. The following sections describe in details the ellipse detection algorithm and the two segmentation algorithms used to divide the stroke. % The input to these algorithms are the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing. The output is a set of dominant points which are connected by either lines or curves. 
%After two different segmentations are generated, 
%///AllBlockDiagram.jpg  you can change the block to the main block. flowchart
 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{images/flowchart.jpg}
	\caption{Segmentation process flow chart} Inputs to the segmentation proces are the input stroke and \textit{possible dominant points} $P_{pd}$. The output is either the stroke fitted as an ellipse or the best segmentation generated from the two segmentation algorithms.  % First, attempts to fit stroke as an ellipse. Then two different segmentation are generated using (AlgS1 and AlgS2). Finaly, the Minimum error of segmentation as the best segmentation. 
	\label{fig:segblock}
\end{figure}

% \begin{figure}
%	\centering
%		\includegraphics[scale=0.48]{images/blockSmall.jpg}
%	\caption{Segmentation Block} 
%	\label{fig:segblock}
%\end{figure}

\subsection{Ellipse fitting}
%The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov}.  
This process attempts to fit the stroke points into an ellipse arc; it starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axis of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ and $height$ are the width and height of the bonding box. The least square fitting algorithm \cite{chernov} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})  
\begin{equation}
E = \sum\limits_{i = 0}^N {\frac{{(x_i - x_0 )}}{{a^2 }}^2  + \frac{{(y_i - y_0 )}}{{b^2 }}^2  - 1} 
\label{eq:circleFit}
\end{equation}
where $N$ is number of points in the stroke, $a$ \& $b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point and $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke. A list of new values for $x_0$ , $y_0$ ,$a$ and $b$ are generated randomly from the older values with small increments after each loop.  After few iterations, the final fit error of the estimated ellipse is reported. Another measure is used to compute the efficiency of the final estimated ellipse. The value of $eff$ in Equation \ref{eq:circleError} ensures that the drawn percentage of ellipse is considered. This eliminates fitting a line into very large ellipse but leaves small ellipses to be fitted as a partial or a full ellipse.

 \begin{equation}
%eff= (P_{percent}/E)
%eff = \left\{ {\begin{array}{*{20}c}
%   {\frac{{P_{percent} }}{E}} & {ifP_{percent}  < 0.5}  \\
%   {\frac{{P_{percent} }}{{E \times 2}}} & {ifP_{percent}  \ge 0.5}  \\
%\end{array}} \right,
eff= \begin{cases} 
\displaystyle \frac{P_{percent}}{E},&if P_{percent}< 0.5 \\ \medskip
\displaystyle  \frac{P_{percent}}{2\times E},&if P_{percent}\geq 0.5 \\
% \frac{P_{percent}}{2\times E},&\mbox{if} P_{percent}\mbox{\geq0.5} 
\end{cases},
\label{eq:circleError}
\end{equation}
 where $E$ is the error computed by Equation (\ref{eq:circleFit}), and 
\[
P_{percent}  = L_{stroke} /P_{ellipse}, 
%\label{eq:ErrorArea}
\]
 where $L_{stroke}$ total length of stroke and $P_{ellipse} $ is the perimeter of the estimated ellipse. If $eff$ has a value more than threshold $th_{Ellipse}$\footnote{By trial and error best threshold found was $th_{Ellipse}=0.2$} then the stroke is segmented as an ellipse otherwise the system proceeds to the next step.

\subsection{Non ellipse fitting algorithm}
\label{subsubsec:Discreteparticleswarmalgorithm}
\begin{figure}
	\centering
		%\subfigure[Possible dominate point]{ 	\label{fig:LabelsPPD}	\includegraphics[scale=0.7]{images/ppd.jpg}}
	%	\hfill
	 \includegraphics[scale=0.7]{images/pso1.jpg}			
	\caption{DPSO algorithm encoding} The swarm particle encoding where 1 represent the dominant points or segment points (i.e points are drawn as a big plus). Length L is the distance between a stroke point and the line connecting two successive segment points. %represent the length   % a) Possible dominate point b) Particle encoding  
	\label{fig:pso1}
\end{figure}
Two DPSO algorithms are used to generate different segmentations for each stroke the user draws. For each segmentation generated, an error is evaluated. The segmentation with the minimum error value is chosen as the best stroke segmentation (Figure \ref{fig:segblock}). The problem definition is the same in both algorithms but they differ in the method they use to compute fitness and error functions. 
\begin{description}
	\item[ Problem definition:] The input stroke $S$  with $N$ points can be represented by set $S = \left\{ {x_1 ,x_2  \ldots x_N }\right\}$ where $x_i$ is the location of the point $i$. The swarm algorithms consist of $M$ agents which are represented by the set $A = \left\{ {P_i \left| {i = 1,2 \cdots M} \right.} \right\}$ where $P_i$ is a single solution particle from the solution space. Each particle decodes the problem using a binary array with the same length $N$ as the input stroke.

@@

After each iteration of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedures: 
\begin{enumerate}
	\item For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled, the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. 
	\item The particles are tested to make sure that the distance between every two successive dominate point is larger than $min_D$, where $min_D$ is 5\% of the total length of the stroke.  Otherwise, two temperary particles are evaluated by removing either point from the current particle. After that, one of these points is removed based on the error caused by its removal.
	\item Each two successive segments are tested to determine if they can be merged into a single segment. If merging the two segments results in segmentation that has an error value less than or equal to the error before merging, the two segments are merged. This step ensures that the final segmentation of the system has minimum number of segments. 
	% the same or smaller the error
\end{enumerate}
%1) 2)  % If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features (FS1)}
  They are features that define the structure of the geometrical symbol such as;   
 \begin{itemize}
	 \item \emph{Segments:} The number of segments in the symbol.
	 \item \emph{Strokes:} The number of strokes or partial strokes that created the symbols.  
		\item  \emph{Primitives:} The number of primitives in the symbol. This feature helps when identifying  symbols with mixed geometric primitives such as cylinders and callouts.  
		\item \emph{Curves:} The number of curves or ellipses in the symbol. 
		\item \emph{Lines:} The number of lines in the symbol. 
		\item \emph{Perpendicular}\emph{lines:} The number of perpendicular lines. 
		\item \emph{Parallel}\emph{lines:} The number of parallel lines. 
		\item \emph{Intersections:} The number of intersection between lines and curves. 
		\item \emph{T intersections:} The number of T intersections. 
		\item \emph{L intersections:} The number of L intersections. 
		\item \emph{X intersections:} The number of X intersections.	
		%\item[No. of holes] Number of holes in the symbols. 
\end{itemize}
\subsubsection{Rubine Features (FS2)}
  These features were introduced by Rubine \cite{gestureexample12} for single stroke gestures. Such as length of stroke, cosine of starting angle, smoothness of symbol and some features extracted from the bounding box of the symbol.

@@

 \begin{figure}
			\subfigure[The Digital Design Symbols Dataset(LD-DS)]{\label{fig:LogicsymbolSet} \includegraphics[scale=0.5]{images/logicSet.jpg}}
			\hfill
			\subfigure[The Electrical Symbols Dataset(EL-DS)]
			 {\label{fig:ELsymbolSet}\includegraphics[scale=0.45]{images/EelectImage.jpg}}
%			 \label{fig:ELsymbolSet} 
%			 \fbox{ 
%			 \parbox{4cm}{\includegraphics[scale=0.5]{images/EelectImage.jpg}	}}}

@@

 \begin{figure}
	\centering		
	 	\subfigure[Segmentation Error Vs. Swarm Iterations]{	\label{fig:swarmtesting} \includegraphics[scale=0.3]{images/swarmIterationVsParticles.jpg}}
	 %\includegraphics[scale=0.3]{images/swarmAgentTest.jpg}} 	\caption{}

	%\caption{Experiments results :} a)    b) c) 
	%\end{figure}

%\begin{figure}
	%\centering

@@

	%\Segmentation Error Vs. PSO parameters. caption{Experiments results :} a)    b) c)

@@

% \begin{table*}
% 	\centering
% 	%\small
% 		%	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}\hline 
%  \include{conf2}
%  		%\end{tabular}
%  	%}
% 	\caption{Confusion Matrix of DL-DS}
% 	\label{tab:ConfusionMatrix}
% 	%\end{minipage}
% \end{table*}

@@

 \begin{figure*}[h]
	\centering
	\subfigure[Feature Comparison on Hs-DS]{	\label{fig:testFeaturesAll} \includegraphics[scale=0.5]{images/featuresdifference.jpg}}
	\subfigure[Feature Comparison on EL-DS]{	\label{fig:ELtestFeaturesAll}	\includegraphics[scale=0.34]{images/FeatEl.jpg}}
	\caption{Different Features} The effect of different feature set on accuracy.  FS4 and FS3 give good results in a) Hs-DS and b) EL-Ds. The results are improved by combining FS1 with FS2 and FS1 with FS4.    %The graph shows the recognition rate of each symbol using different algorithms. 
\end{figure*}

@@

 The main features affected by the segmentation stage are the geometric features computed in \textbf{FS1}. Hence those are the features we used to estimate the efficiency of the segmentation algorithm. The recognition accuracy using only the geometrical features is reported in Figure \ref{fig:testFeatonly}. The figure shows that \textsl{AlgS2} achieve better results.  %effect of the segmentation algorithm 
 %Figure \ref{fig:SampleSeg} shows a sample of the output of the segmentation block. The result of segmentation block is tested by their effect on the final recognition results using the features that will be highly affected using different algorithms like \textbf{FS1} and \textbf{FS4} (Section\ref{sec:Recognition}). Figure \ref{fig:testFeatonly} shows the result of the different algorithms using only \textbf{FS1} features. 
\begin{figure*} 
	\centering
		\includegraphics[scale=0.3]{images/OnlyF1EL.jpg}
	\caption{The effect of segmentation on accuracy} The experiment is implemented on Hs-Ds using only FS1.  %The effect of symbol complexity.  %The graph shows the recognition rate of each symbol using different algorithms. 
	\label{fig:testFeatonly}
\end{figure*}

@@

	\caption{Segmentation outputs} The output segmentation of different input strokes. a) Sample of single input stroke b) The single stroke after segmentation. c) An example of multiple strokes and their segmentation. %The figure shows sample of the output of the segmentation phase. Figure a) shows the stroke after the user draws it. The red points are points labeled as \textit{possible dominate points}. Figure b) shows the symbol after being segmented. Figure c) shows the segmentation of different and overlapped strokes. Notice the system mark the center of the ellipse that was identified.
	\label{fig:SampleSeg}
\end{figure}

>>>>

Recognition based on AI methods as applied in \cite{SketchRead2007} has the principle problem of computational complexity of the matching between shape description and geometric primitives produced in the segmentation process. Natural sketch are difficult to be segmented and grouped to create reliable fragment of geometric primitives. This is partially due to the increase of noise and other phenomena such as overlapped, over traced strokes and sharing of strokes between different symbols. Since these systems are dependable on the segmentation results, they either perform recognition using undependable segmentation results or consider different ways to segment the input sketch. This is because they assume that each stroke plays a role in the structure of the shape.

@@

 The remaining of the paper is as follows; the literature review on stroke segmentation is presented in Section \ref{sec:review}. Section \ref{sec:ParticleSwarmAlgorithm} introduces Particle Swarm Optimization. Section \ref{Sysdisc} describes the proposed system in details. Section \ref{sec:Experiments} describes the results obtained from the experiments that were preformed. Finally, the conclusion and future work is presented in Section \ref{ConclusionandFutureWork}.

@@

\begin{equation}
%\[
p_{ij}=p_{ij}+v_{ij},
%\
\label{eq:Swarm1}
\end{equation}

where $p_{ij}$ represent the $j$th dimension in the $i$th particle and $v_{ij}$ is the velocity of the $j$th dimension in the $i$th particle.
 %Equation [\ref{eq:Swarm}] shows how velocity and direction of each particle are computed
 \begin{equation}
v_{ij}  = v_{ij} \omega + c_1 r_1 (lbest_{ij}  - p_{ij} ) + c_2 r_2 (gbest_{ij}  - p_{ij} ),
\label{eq:Swarm}
\end{equation}
 where $\omega$ is the inertia weight parameter which controls the tradeoff between exploration and exploitation,  $lbest_{ij}$ is the local best particle, $gbest_{ij}$ is the global best particle, $r_1$ \& $r_2$ are random variables and $c_1$ \& $c_2$ are the swarm acceleration parameters.

@@

\subsection{Preprocessing}
\label{Prepross} 
 Speed and time difference information were widely used in sketch recognition systems \cite{earlyprocess}. Figure \ref{fig:orignalStroke} shows an example of an input stroke drawn by users.  Figure \ref{fig:speed2Distance} shows the graph of time difference and the speed for the stroke in Figure \ref{fig:orignalStroke}. It was noted that time difference provides more distinct maxima than the speed information. Agar et al. \cite{polygonfeedback31} mentions that the pointing device (e.g., mouse) sampling rate is the reason for this phenomenon. The points are sampled at regular time intervals as long as the pointing device is moving. Therefore, there are no new detected points when the pointing device is stationary. This leads to a nearly constant time difference between samples while the pen is moving and large difference while the pen is stationary (Figure \ref{fig:timediff}). Contrary to time difference information, the speed information has a lot of noise in the data because usually  users draw with variant speeds (Figure \ref{fig:speed2Distance}).  %Similarly, direction information is used as it provides better distinctive maxima for the corners than the curvature information \cite{meanshift10}.% Figure \ref{fig:curvatures} shows the direction and curvature graphs for the stroke drawn in Figure \ref{fig:orignalStroke}

@@

 \begin{figure}
	\centering
			\subfigure[Speed versus Displacement Graph]{	\includegraphics[scale=0.4]{images/vel3.jpg}}
			\hfill
			\subfigure[ Time Difference  versus Displacement Graph] {\label{fig:timediff}\includegraphics[scale=0.4]{images/td3.jpg}}
	\caption{Speed and Time Difference Graphs}  a) speed graph has lot of noise in the data with respect to b) time difference.   The black points denote the location of the possible dominant points $P_{pd}$. Black rectangles represent regions $R_i$ that in a) Are lower than the threshold. b) Are above the threshold. 
	\label{fig:speed2Distance}
\end{figure}

Dominant points are characterized by low speed values, high curvature and direction values. However, the dominant points from the curvature information may not be the same points as the dominant points extracted from the speed information. Hence, we compute time difference, direction, speed and curvature of each point along a stroke to ensure that no dominant point is overlooked in this phase. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ calculated as the angle between vectors $\overrightarrow {P_{i - 1} P_i }$ and the $x-axis$ and the curvature is considered as the change in direction $d$ with respect to length $s$ i.e. $c= \Delta d/\Delta s$.

@@

\subsection{Segmentation}
\label{seg}
In the segmentation stage, the input stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{ellipsefit}. The ellipse fitting step mainly helps to prevent the system from over segmenting the stroke into multiple lines or curves if the input stroke is an ellipse. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two further segmentation algorithms that divide the stroke to either lines or lines and curves. Hence we generate two different segmentations; the system chooses the segmentation with the minimum error. The following sections describe in details the ellipse detection algorithm and the two segmentation algorithms used to divide the stroke. 
 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{images/flowchart.jpg}
	\caption{Segmentation process flow chart} Inputs to the segmentation proces are the input stroke and \textit{possible dominant points} $P_{pd}$. The output is either the stroke fitted as an ellipse or the best segmentation generated from the two segmentation algorithms.   
	\label{fig:segblock}
\end{figure}

\subsection{Ellipse fitting}

This process attempts to fit the stroke points into an ellipse arc; it starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axis of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ and $height$ are the width and height of the bonding box. The least square fitting algorithm \cite{chernov} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})  
\begin{equation}
E = \sum\limits_{i = 0}^N {\frac{{(x_i - x_0 )}}{{a^2 }}^2  + \frac{{(y_i - y_0 )}}{{b^2 }}^2  - 1} 
\label{eq:circleFit}
\end{equation}
where $N$ is number of points in the stroke, $a$ \& $b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point and $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke. A list of new values for $x_0$ , $y_0$ ,$a$ and $b$ are generated randomly from the older values with small increments after each loop.  After few iterations, the final fit error of the estimated ellipse is reported. Another measure is used to compute the efficiency of the final estimated ellipse. The value of $eff$ in Equation \ref{eq:circleError} ensures that the drawn percentage of ellipse is considered. This eliminates fitting a line into very large ellipse but leaves small ellipses to be fitted as a partial or a full ellipse.

 \begin{equation}
 eff= \begin{cases} 
\displaystyle \frac{P_{percent}}{E},&if P_{percent}< 0.5 \\ \medskip
\displaystyle  \frac{P_{percent}}{2\times E},&if P_{percent}\geq 0.5 \\

\end{cases},
\label{eq:circleError}
\end{equation}
 where $E$ is the error computed by Equation (\ref{eq:circleFit}), and 
\[
P_{percent}  = L_{stroke} /P_{ellipse}, 
%\label{eq:ErrorArea}
\]
 where $L_{stroke}$ total length of stroke and $P_{ellipse} $ is the perimeter of the estimated ellipse. If $eff$ has a value more than threshold $th_{Ellipse}$\footnote{By trial and error best threshold found was $th_{Ellipse}=0.2$} then the stroke is segmented as an ellipse otherwise the system proceeds to the next step.

\subsection{Non ellipse fitting algorithm}
\label{subsubsec:Discreteparticleswarmalgorithm}
\begin{figure}
	\centering
 	 \includegraphics[scale=0.7]{images/pso1.jpg}			
	\caption{DPSO algorithm encoding} The swarm particle encoding where 1 represent the dominant points or segment points (i.e points are drawn as a big plus). Length L is the distance between a stroke point and the line connecting two successive segment points.  
	\label{fig:pso1}
\end{figure}
Two DPSO algorithms are used to generate different segmentations for each stroke the user draws. For each segmentation generated, an error is evaluated. The segmentation with the minimum error value is chosen as the best stroke segmentation (Figure \ref{fig:segblock}). The problem definition is the same in both algorithms but they differ in the method they use to compute fitness and error functions. 
\begin{description}
	\item[ Problem definition:] The input stroke $S$  with $N$ points can be represented by set $S = \left\{ {x_1 ,x_2  \ldots x_N }\right\}$ where $x_i$ is the location of the point $i$. The swarm algorithms consist of $M$ agents which are represented by the set $A = \left\{ {P_i \left| {i = 1,2 \cdots M} \right.} \right\}$ where $P_i$ is a single solution particle from the solution space. Each particle decodes the problem using a binary array with the same length $N$ as the input stroke.

@@

After each iteration of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedures: 
\begin{enumerate}
	\item For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled, the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. 
	\item The particles are tested to make sure that the distance between every two successive dominate point is larger than $min_D$, where $min_D$ is 5\% of the total length of the stroke.  Otherwise, two temperary particles are evaluated by removing either point from the current particle. After that, one of these points is removed based on the error caused by its removal.
	\item Each two successive segments are tested to determine if they can be merged into a single segment. If merging the two segments results in segmentation that has an error value less than or equal to the error before merging, the two segments are merged. This step ensures that the final segmentation of the system has minimum number of segments. 
	% the same or smaller the error
\end{enumerate}
%1) 2)  % If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features (FS1)}
  They are features that define the structure of the geometrical symbol such as;   
 \begin{itemize}
	 \item \emph{Segments:} The number of segments in the symbol.
	 \item \emph{Strokes:} The number of strokes or partial strokes that created the symbols.  
		\item  \emph{Primitives:} The number of primitives in the symbol. This feature helps when identifying  symbols with mixed geometric primitives such as cylinders and callouts.  
		\item \emph{Curves:} The number of curves or ellipses in the symbol. 
		\item \emph{Lines:} The number of lines in the symbol. 
		\item \emph{Perpendicular}\emph{lines:} The number of perpendicular lines. 
		\item \emph{Parallel}\emph{lines:} The number of parallel lines. 
		\item \emph{Intersections:} The number of intersection between lines and curves. 
		\item \emph{T intersections:} The number of T intersections. 
		\item \emph{L intersections:} The number of L intersections. 
		\item \emph{X intersections:} The number of X intersections.

\end{itemize}
\subsubsection{Rubine Features (FS2)}
  These features were introduced by Rubine \cite{gestureexample12} for single stroke gestures. Such as length of stroke, cosine of starting angle, smoothness of symbol and some features extracted from the bounding box of the symbol.

@@

 \begin{figure}
			\subfigure[The Digital Design Symbols Dataset(LD-DS)]{\label{fig:LogicsymbolSet} \includegraphics[scale=0.5]{images/logicSet.jpg}}
			\hfill
			\subfigure[The Electrical Symbols Dataset(EL-DS)]
			 {\label{fig:ELsymbolSet}\includegraphics[scale=0.45]{images/EelectImage.jpg}}

@@

 \begin{figure}
	\centering		
	 	\subfigure[Segmentation Error Vs. Swarm Iterations]{	\label{fig:swarmtesting} \includegraphics[scale=0.3]{images/swarmIterationVsParticles.jpg}}

@@

@@

@@

 \begin{figure*}
	\centering
	\subfigure[Feature Comparison on Hs-DS]{	\label{fig:testFeaturesAll} \includegraphics[scale=0.5]{images/featuresdifference.jpg}}
	\subfigure[Feature Comparison on EL-DS]{	\label{fig:ELtestFeaturesAll}	\includegraphics[scale=0.34]{images/FeatEl.jpg}}
	\caption{Different Features} The effect of different feature set on accuracy.  FS4 and FS3 give good results in a) Hs-DS and b) EL-Ds. The results are improved by combining FS1 with FS2 and FS1 with FS4.    %The graph shows the recognition rate of each symbol using different algorithms. 
\end{figure*}

@@

 The main features affected by the segmentation stage are the geometric features computed in \textbf{FS1}. Hence those are the features we used to estimate the efficiency of the segmentation algorithm. The recognition accuracy using only the geometrical features is reported in Figure \ref{fig:testFeatonly}. The figure shows that \textsl{AlgS2} achieve better results.  
\begin{figure*} 
	\centering
		\includegraphics[scale=0.3]{images/OnlyF1EL.jpg}
	\caption{The effect of segmentation on accuracy} The experiment is implemented on Hs-Ds using only FS1.  %The effect of symbol complexity.  %The graph shows the recognition rate of each symbol using different algorithms. 
	\label{fig:testFeatonly}
\end{figure*}

@@

	\caption{Segmentation outputs} The output segmentation of different input strokes. a) Sample of single input stroke b) The single stroke after segmentation. c) An example of multiple strokes and their segmentation.   identified.
	\label{fig:SampleSeg}
\end{figure}
